Ornaments as indicators of social changes in northeastern Taiwan before and after the European colonial period

### 2020-04-19 analysis/paper/paper.Rmd
edits according to R3 comments

The archaeological indicators of the start of the European influence at Kiwulan are the appearance of light grey glazed jars, known as "An-ping" jars in China and Taiwan, and large dark brown glazed stoneware jars that were introduced to Taiwan during the early 17th century. Large dark brown glazed stoneware jars may have been made in Southeast Asia, and are frequently found in European shipwrecks from this period as vessels for transporting water, wine or other liquids on long voyages. The earliest evidence of light gray glazed jars in this region has been found among the cargo of the Spanish shipwreck _San Diego_, which sunk in 1600 AD [@Hsieh2009; @Hsieh1995]. Southeast China is assumed to be the origin of the light gray glazed jars, however these are commonly found at sites in Taiwan that were associated with European activities, such as the Zeelandia fort site in Tainan [@Wang2007]. The jar shapes found at Kiwulan are typical of those found elsewhere in VOC sites occupied during the 17th century [@Berrocal2018, pp.917; @Cort2017, pp.282; @Grave2013; @Ketel2011; @Klose2018, pp.131]. We cannot be sure of the exact process that bought them to Kiwulan: they might have been directly imported by Europeans, by Chinese merchants, or by indigenous groups via regional networks in north Taiwan. In any case, the high volume of ceramics transported by Europeans, and their high mobility in shipping trade played an important role in introducing foreign jars to Taiwan. Those jars were widely distributed across the site and can serve as indicators, together with the radiocarbon dates, to identify the excavation units associated with the pre-European period and the start of European influence at Kiwulan. In addition to stoneware jars as indicators of European presence, around 300 pieces of locally made pipes and a few imported pipes were found at Kiwulan. Smoking is likely to have been introduced by Europeans, and we found that the presence of pipes in the archaeological record here is consistent with distributions of glazed jar fragments, which are far more numerous and widespread across the site (n = 1685). Those temporal diagnostic ceramics might have been continually used in later period that could mislead our chronology. Thus, we focus on identifying the certain context with high frequency of those ceramics that can reflect the early 17th century when they prevailed.

@@

table_kw_dates <-
  kw_dates %>% 
  mutate(`Lab code` = str_remove(`Lab code`, " "),
         `Uncalibrated Age BP` = `Date BP`) %>% 
  mutate(`Above mean sea level (cm)` = paste0(Depth1, " to ", Depth2)) %>%
  mutate(`Pit-Layer` = paste0(Square, "-", Layer)) %>% 
  left_join(kw_age_cali_df_tidy, "Pit-Layer") %>% 
  mutate(ave = (`2.5%` + `97.5%`)/ 2) %>% 
  mutate(ave = ifelse(is.na(ave), 200, ave)) %>%  
  arrange(ave) %>% 
  mutate(`Calibrated Age BP (95% credible interval)` = 
           ifelse(is.na(`Calibrated date (95% conf. level)`), 
                  "", 
                  `Calibrated date (95% conf. level)`)) %>% 
  mutate(`Uncalibrated Age BP` = str_remove_all(`Uncalibrated Age BP`, "yr")) %>% 
  dplyr::select(`Lab code`, 
                `Pit-Layer`, 
                `Depth (cm)`, 
                `Uncalibrated Age BP`, 
                `Calibrated Age BP (95% credible interval)`, 
                Context)

@@

Ornaments (Figure \@ref(fig:ornaments-image)) were found in a variety of archaeological contexts including post-holes area, burials, and middens. This study focuses on `r sam_post_hole` ornaments from 40 sampling squares located at the dwelling place of Kiwulan, indicated by aligned post-holes with _in-situ_ posts (Figure \@ref(fig:KWL-sam-area)). Occupation floors were not identified during excavation. We choose these units because they were stratigraphically intact and undisturbed by modern construction activity, compared to excavation squares on the periphery of the site. There are 35 burials in the sampling area, one third of the total number of burials at Kiwulan. Ornaments are commonly used as grave goods in burials, with the total number of ornaments in burials including 3,173. Besides, there are 27 ornaments found in midden contexts. The obvious difference in number in burials and midden contexts indicates the value of ornaments as prestige goods. The high number in burials is also due to the presence of bead strands or patterned bands of beads, which sometimes contains thousands of beads in an individual burial [@Chen2007]. The unequal distribution of beads between burials was interpreted as the evidence for hierarchy by @Cheng2008. However, @Hsieh2012 suggests a relatively egalitarian form from the comparative analysis of frequency and proportion of burial goods. To have a comprehensive understanding, we focus on ornaments from the dwelling place contexts (Figure \@ref(fig:plot-all-ornaments-count), Table \@ref(tab:tidy-data-for-table)) because these give us the greatest spatial and temporal representation across the three time periods, and so are most informative of social inequality as indicated by uneven distributions of ornaments. The burials are excluded because most burials from the sampling area date to the European period (n = 21), limiting the usefulness of comparisons between the periods. Also, it is beyond the scope of this research. Ornaments could be potential heirlooms inherited over multiple generations. However, there is no continuous increase over time that suggests accumulation and discard of ornaments is not constant. We assume that changes in the abundance of ornaments correspond to changes in the amount we see in archaeological record.

```{r, ornaments-image, fig.cap = "Subtypes of ornament in each major class. A: carnelian beads, B: bells, C: glass beads and gold-foil beads (far right), D: metal rings. Photographs are presented in the same order as those subtypes in the table but from left to right instead. The photographs of B, C, D classes were from original excavation report (Chen 2007)", echo=FALSE}
knitr::include_graphics(here("analysis", "figures", "ornament.png"))
```

@@

```{r metal-fre-plot}
# plot for different shape of metal ring by period
ring_shape <-
  ornaments_shape_count %>% 
  filter(Categories == 'Metal ring',
         !Type %in% c('large thick string',
                       'small thin string')) %>% 
  ungroup() %>% 
  mutate(Type = fct_drop(Type)) %>% 
  mutate(Type = fct_infreq(Type)) %>% 
  mutate(Type = fct_rev(Type)) 
ring_shape_plot <-
ggplot(ring_shape) +
  geom_col(aes(Type, n)) +
  coord_flip() +
  labs(y = "",
       x = "Metal ring") +
  facet_wrap(~ period) +
  theme_minimal(base_size = base_size_multi_plot) 
```

@@

An indirect colonial influence may be indicated at Kiwulan by the greater diversity of ornament types and materials during the European period. Yilan was involved in complex trading networks both on a regional scale with other indigenous groups and Chinese merchants, and at a global scale with Europeans, including the Dutch and the Spanish. The ornaments have multiple origins, including Southeast Asia and China, and were first introduced into northeastern Taiwan by Chinese merchants before 17th century. Later, trade activities became more frequent and intense in the 17th century due to European activities. The greater diversity and quantity of ornaments likely results from participation in large scale exchange networks that stimulated the circulation of different ornament classes. The frequency of overall ornaments and each subtype declines significantly after European influence fades during the Chinese period in the early 19th century. This may be due to a smaller scale of trading networks, the overall decline of indigenous populations in Yilan, or the adoption of Han Chinese practices. The decline of the population at Kiwulan may be related to the movement of many indigenous people southwards to Hualien due to increasing Han Chinese immigrants who took over their lands at the end of the 18th century [@Chen2007].

@@

How might these results fit into a bigger picture of social change at periphery of colonial systems? We may get some insight into the general pathways that led to social inequality in northeastern Taiwan by considering how people have achieved and maintained power in a wide variety of societies [@Ames2010; @Bowles2010; @Drennan2010; @Feinman2000;]. The corporate/network model proposed by @Feinman2000 expands traditional hierarchical complexity to provide a comparative basis for distinct strategies for power. In the network mode, inequality develops when individuals accumulate wealth through their individual networks and people use their wealth to attract factions, control resources, and monopolize trade networks. In contrast, the corporate mode stresses shared power across different groups and sectors, integrative ceremonies and rituals, and large cooperative labour tasks [@Feinman2000; @Siegel1999]. We may be able to interpret Yilan social organization as moving from corporate mode before the European arrival, to a network mode during European presence. The small number of ornaments, and less concentrated distribution during the pre-European period appears consistent with shared power and wealth of the corporate mode. The long-distance trade network introduced by Europeans resulted in the appearance of a network mode, and the emergence of competition among ambitious individuals for prestige, wealth, or power through collecting trade goods [@Brumfiel1994; @Clark1994]. Because of the weak direct control from the European colonizers in northeastern Taiwan, local leaders may have had the flexibility to manipulate European colonial images, expand personal power, and monopolize the high-value trade goods [@Kang2012].

That said, the evidence from Kiwulan may be consistent with a variety of scenarios of indigenous-colonial relations. The increasing number and concentrated spatial patterns of ornaments may indicate a practice of cultural resistance against the European intrusion. Resistance to European economic and political demands may be inferred if ornaments were used as a display of social identity and to emphasize the local customs that had existed before European contact [cf. @Rubertone2000]. Resistance could be presented in many forms. Another scenario is that ornaments were treated as heirlooms, such as carnelian beads and gold-foil beads, that passed from one generation to the next, accumulating at Kiwulan over time. This process would result in a natural increase in ornaments over time, unrelated to colonial influences, but has limited value in explaining the shifts in spatial patterns.

>>>>

The archaeological indicators of the start of the European influence at Kiwulan are the appearance of light grey glazed jars, known as "An-ping" jars in China and Taiwan, and large dark brown glazed stoneware jars that were introduced to Taiwan during the early 17th century. Large dark brown glazed stoneware jars may have been made in Southeast Asia, and are frequently found in European shipwrecks from this period as vessels for transporting water, wine or other liquids on long voyages. The earliest evidence of light gray glazed jars in this region has been found among the cargo of the Spanish shipwreck _San Diego_, which sunk in 1600 AD [@Hsieh2009; @Hsieh1995]. Southeast China is assumed to be the origin of the light gray glazed jars, however these are commonly found at sites in Taiwan that were associated with European activities, such as the Zeelandia fort site in Tainan [@Wang2007]. The jar shapes found at Kiwulan are typical of those found elsewhere in VOC sites occupied during the 17th century [@Berrocal2018, pp.917; @Cort2017, pp.282; @Grave2013; @Ketel2011; @Klose2018, pp.131]. We cannot be sure of the exact process that brought them to Kiwulan: they might have been directly imported by Europeans, by Chinese merchants, or by indigenous groups via regional networks in north Taiwan. In any case, the high volume of ceramics transported by Europeans, and their high mobility in shipping trade played an important role in introducing foreign jars to Taiwan. Those jars were widely distributed across the site and can serve as indicators, together with the radiocarbon dates, to identify the excavation units associated with the pre-European period and the start of European influence at Kiwulan. In addition to stoneware jars as indicators of European presence, around 300 pieces of locally made pipes and a few imported pipes were found at Kiwulan. Smoking is likely to have been introduced by Europeans, and we found that the presence of pipes in the archaeological record here is consistent with distributions of glazed jar fragments, which are far more numerous and widespread across the site (n = 1685). Those temporal diagnostic ceramics might have been continually used in later period that could mislead our chronology. Thus, we focus on identifying the certain context with high frequency of those ceramics that can reflect the early 17th century when they prevailed.

@@

table_kw_dates <-
  kw_dates %>% 
  mutate(`Lab code` = str_remove(`Lab code`, " "),
         `Uncalibrated Age BP` = `Date BP`) %>% 
  mutate(`Depth (cm)` = paste0(Depth1, " to ", Depth2)) %>%
  mutate(`Pit-Layer` = paste0(Square, "-", Layer)) %>% 
  left_join(kw_age_cali_df_tidy, "Pit-Layer") %>% 
  mutate(ave = (`2.5%` + `97.5%`)/ 2) %>% 
  mutate(ave = ifelse(is.na(ave), 200, ave)) %>%  
  arrange(ave) %>% 
  mutate(`Calibrated Age BP (95% credible interval)` = 
           ifelse(is.na(`Calibrated date (95% conf. level)`), 
                  "", 
                  `Calibrated date (95% conf. level)`)) %>% 
  mutate(`Uncalibrated Age BP` = str_remove_all(`Uncalibrated Age BP`, "yr")) %>% 
  dplyr::select(`Lab code`, 
                `Pit-Layer`, 
                `Depth (cm)`, 
                `Uncalibrated Age BP`, 
                `Calibrated Age BP (95% credible interval)`, 
                Context)

@@

Ornaments (Figure \@ref(fig:ornaments-image)) were found in a variety of archaeological contexts including post-holes area, burials, and middens. For the ornament categories, we follow the well established topology based on raw materials and shapes for the region of northeastern Taiwan [@Chen2007; @NMTH2005]. This study focuses on `r sam_post_hole` ornaments from 40 sampling squares located at the dwelling place of Kiwulan, indicated by aligned post-holes with _in-situ_ posts (Figure \@ref(fig:KWL-sam-area)). Occupation floors were not identified during excavation. We choose these units because they were stratigraphically intact and undisturbed by modern construction activity, compared to excavation squares on the periphery of the site. There are 35 burials in the sampling area, one third of the total number of burials at Kiwulan. Ornaments are commonly used as grave goods in burials, with the total number of ornaments in burials including 3,173. Besides, there are 27 ornaments found in midden contexts. The obvious difference in number in burials and midden contexts indicates the property of ornaments as prestige goods. The high number in burials is due to the presence of bead strands or patterned bands of beads, which sometimes contains thousands of beads in an individual burial [@Chen2007]. The unequal distribution of beads between burials was interpreted as the evidence for hierarchy by @Cheng2008. However, @Hsieh2012 suggests a relatively egalitarian form from the comparative analysis of frequency and proportion of burial goods. To have a comprehensive understanding, we focus on ornaments from the dwelling place contexts (Figure \@ref(fig:plot-all-ornaments-count), Table \@ref(tab:tidy-data-for-table)) because these give us the greatest spatial and temporal representation across the three time periods, and so are most informative of social inequality as indicated by uneven distributions of ornaments. The burials are excluded because most burials from the sampling area date to the European period (n = 21), limiting the usefulness of comparisons between the periods. Ornaments could be potential heirlooms inherited over multiple generations. However, there is no continuous increase over time that suggests accumulation and discard of ornaments is not constant. Thus, we assume that changes in the abundance of ornaments reflect changes in the amount in archaeological record. Houses and burials may also be a useful source of evidence to understand population size but a proper treatment of those is beyond the scope of the paper.

```{r, ornaments-image, fig.cap = "Subtypes of ornament in each major class. A: carnelian beads, B: bells, C: glass beads (the far left and the middle) and gold-foil beads (the far right), D: metal rings. Photographs are presented in the same order as those subtypes in the table but from left to right instead. The photographs of B, C, D classes were from original excavation report (Chen 2007)", echo=FALSE}
knitr::include_graphics(here("analysis", "figures", "ornament.png"))
```

@@

```{r metal-fre-plot}
# plot for different shape of metal ring by period
ring_shape <-
  ornaments_shape_count %>% 
  filter(Categories == 'Metal ring',
         !Type %in% c('large thick string',
                       'small thin string')) %>% 
  ungroup() %>% 
  mutate(Type = fct_drop(Type)) %>% 
  mutate(Type = fct_infreq(Type)) %>% 
  mutate(Type = fct_rev(Type)) 
ring_shape_plot <-
ggplot(ring_shape) +
  geom_col(aes(Type, n)) +
  coord_flip() +
  labs(y = "",
       x = "Metal ring") +
  facet_wrap(~ period) +
  scale_y_continuous(breaks=c(0,2,4,6,8)) +
  theme_minimal(base_size = base_size_multi_plot) 
```

@@

An indirect colonial influence may be indicated at Kiwulan by the greater diversity of ornament types and materials during the European period. Yilan was involved in complex trading networks both on a regional scale with other indigenous groups and Chinese merchants, and at a global scale with Europeans, including the Dutch and the Spanish. The ornaments have multiple origins, including Southeast Asia and China, and were first introduced into northeastern Taiwan by Chinese merchants before 17th century. Later, trade activities became more frequent and intense in the 17th century due to European activities. The greater diversity and quantity of ornaments likely resulted from participation in large scale exchange networks that stimulated the circulation of different ornament classes. The frequency of overall ornaments and each subtype declines significantly after European influence fades during the Chinese period in the early 19th century. This may be due to a smaller scale of trading networks, the overall decline of indigenous populations in Yilan, or the adoption of Han Chinese practices. The decline of the population at Kiwulan may be related to the movement of many indigenous people southwards to Hualien due to the increasing Han Chinese immigrants who took over their lands at the end of the 18th century [@Chen2007].

@@

How might these results fit into a bigger picture of social change at periphery of colonial systems? We may get some insight into the general pathways that led to social inequality in northeastern Taiwan by considering how people have achieved and maintained power in a wide variety of societies [@Ames2010; @Bowles2010; @Drennan2010; @Feinman2000;]. The corporate/network model proposed by @Feinman2000 expands traditional hierarchical complexity to provide a comparative basis for distinct strategies for power. In the network mode, inequality develops when individuals accumulate wealth through their individual networks and people use their wealth to attract factions, control resources, and monopolize trade networks. In contrast, the corporate mode stresses shared power across different groups and sectors, integrative ceremonies and rituals, and large cooperative labour tasks [@Feinman2000; @Siegel1999]. We may be able to interpret Yilan social organization as moving from corporate mode before the European arrival, to a network mode during European presence. The changes from a less concentrated to a high concentrated distribution before and after the presence of Europeans appears consistent with the shift from shared power and wealth to accumulated wealth and monopolization. One possible explanation for the shift could be the long-distance trade network introduced by Europeans. The rarity and the image of colonial power of foreign trade goods resulted in the emergence of competition among ambitious individuals for prestige, wealth, or power through collecting them [@Boone1992; @Brumfiel1994; @Clark1994]. Because of the weak direct control from the European colonizers in northeastern Taiwan, local leaders may have had the flexibility to manipulate European colonial images, expand personal power, and monopolize the high-value trade goods [@Kang2012].

That said, the evidence from Kiwulan may be consistent with a variety of scenarios of indigenous-colonial relations. The increasing number and concentrated spatial patterns of ornaments may also indicate a practice of cultural resistance against the European intrusion. Resistance to European economic and political demands may be inferred if ornaments were used as a display of social identity and to emphasize the local customs that had existed before European contact [cf. @Rubertone2000]. Resistance could be presented in many forms. Another scenario is that ornaments were treated as heirlooms, such as carnelian beads and gold-foil beads, that passed from one generation to the next, accumulating at Kiwulan over time. This process would result in a natural increase in ornaments over time, unrelated to colonial influences, but has limited value in explaining the shifts in spatial patterns.
